---
title: "TCGA-BRCA Genomic Analysis"
subtitle: "Mutation landscape and survival analysis in breast cancer"
author: "Stav"
date: today
output-file: index.html
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  warning: false
  message: false
---

## Overview

This analysis explores somatic mutation patterns and overall survival in the TCGA-BRCA cohort (n = 1,098). Raw clinical, mutation (MAF), and RNA-seq data were downloaded from the NCI GDC and loaded into DuckDB. Staging and mart models were built with dbt before analysis in R.

**Key findings:**

- Stage is the dominant predictor of overall survival (log-rank p < 0.0001)
- TP53 and PIK3CA are the most frequently mutated genes (34% each), consistent with known BRCA biology
- TMB does not stratify survival in univariate analysis but shows a modest independent effect in multivariate Cox regression (HR = 1.035, p = 0.013)
- No individual driver gene mutation significantly predicts survival after adjusting for stage and age

---

```{r setup}
library(duckdb)
library(DBI)
library(dplyr)
library(tibble)
library(ggplot2)
library(survival)
library(survminer)
library(maftools)
library(reactable)
library(broom)
library(forcats)

DB_PATH <- "../data/tcga_brca.duckdb"
con <- dbConnect(duckdb(), dbdir = DB_PATH, read_only = TRUE)

clinical       <- dbGetQuery(con, "SELECT * FROM staging.clinical") %>% tibble()
mutations      <- dbGetQuery(con, "SELECT * FROM staging.mutations") %>% tibble()
patient_summary <- dbGetQuery(con, "SELECT * FROM marts.survival") %>% tibble()
maf_data       <- dbGetQuery(con, "SELECT * FROM raw.maf")

# tertile-based TMB groups
quantiles <- patient_summary %>%
    summarise(
        p33 = quantile(tmb, 0.33, na.rm = TRUE),
        p67 = quantile(tmb, 0.67, na.rm = TRUE)
    )

patient_summary <- patient_summary %>%
    mutate(
        tmb_group = case_when(
            tmb <= quantiles$p33 ~ "low",
            tmb <= quantiles$p67 ~ "medium",
            tmb > quantiles$p67  ~ "high",
            TRUE ~ NA_character_
        ),
        stage_group = factor(stage_group, levels = c("I", "II", "III", "IV"))
    )

maf <- read.maf(maf = maf_data)
```

---

## Data

::: {.panel-tabset}

### Clinical

```{r}
reactable(clinical %>% head(20),
    searchable = TRUE, striped = TRUE, highlight = TRUE, bordered = TRUE)
```

### Mutations

```{r}
reactable(mutations %>% select(patient_id, gene_symbol, variant_classification,
                                variant_type, impact, vaf, is_pass) %>% head(20),
    searchable = TRUE, striped = TRUE, highlight = TRUE, bordered = TRUE)
```

### RNA Expression

```{r}
rna_expression <- dbGetQuery(con, "SELECT * FROM staging.rna_expression LIMIT 20") %>% tibble()
reactable(rna_expression,
    searchable = TRUE, striped = TRUE, highlight = TRUE, bordered = TRUE)
```

:::

---

## Mutation Landscape

### MAF Summary

Overview of variant classifications, types, and SNV classes across the cohort.

```{r maf-summary}
#| fig-height: 8
plotmafSummary(maf, rmOutlier = TRUE, addStat = "median")
```

### Oncoplot — Top 20 Mutated Genes

Each column is a patient, each row is a gene. Colors indicate mutation type. TP53 and PIK3CA dominate at 34% each, consistent with published TCGA-BRCA findings. CDH1 (13%) marks the lobular subtype; GATA3 (13%) marks luminal tumors.

```{r oncoplot}
#| fig-height: 8
oncoplot(maf = maf, top = 20)
```

---

## Survival Analysis

### Stage vs Overall Survival

Kaplan-Meier curves stratified by AJCC pathologic stage. Stage is a highly significant predictor of overall survival (log-rank p < 0.0001), with Stage IV patients showing rapid early mortality.

```{r km-stage}
fit_stage <- survfit(
    Surv(overall_survival_days, vital_status) ~ stage_group,
    data = patient_summary %>% filter(!is.na(stage_group))
)

ggsurvplot(
    fit_stage,
    data    = patient_summary %>% filter(!is.na(stage_group)),
    pval    = TRUE,
    palette = c("#2196F3", "#4CAF50", "#FF9800", "#F44336"),
    legend.title = "Stage",
    legend.labs  = c("I", "II", "III", "IV"),
    xlab = "Days",
    ylab = "Survival Probability"
)
```

### TMB Distribution

Tumor mutation burden is log-normally distributed and centered around 1 mut/Mb — typical for breast cancer, a tumor type with low mutational burden compared to melanoma or lung cancer. The dashed lines show the 5 and 20 mut/Mb thresholds used in literature.

```{r tmb-dist}
patient_summary %>%
    filter(!is.na(tmb)) %>%
    ggplot(aes(x = tmb)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "white") +
    geom_vline(xintercept = c(5, 20), color = "firebrick",
               linetype = "dashed", linewidth = 0.8) +
    scale_x_log10() +
    labs(
        x = "TMB (mutations/Mb, log scale)",
        y = "Count",
        title = "Tumor Mutation Burden Distribution",
        caption = "Dashed lines: 5 and 20 mut/Mb thresholds"
    ) +
    theme_minimal()
```

### TMB vs Overall Survival

KM curves stratified by TMB tertile. Unlike stage, TMB does not significantly separate survival curves (log-rank p = 0.46), consistent with the immunologically cold nature of breast cancer.

```{r km-tmb}
fit_tmb <- survfit(
    Surv(overall_survival_days, vital_status) ~ tmb_group,
    data = patient_summary %>% filter(!is.na(tmb_group))
)

ggsurvplot(
    fit_tmb,
    data    = patient_summary %>% filter(!is.na(tmb_group)),
    pval    = TRUE,
    palette = c("firebrick", "steelblue", "orange"),
    legend.title = "TMB Group",
    legend.labs  = c("High", "Low", "Medium"),
    xlab = "Days",
    ylab = "Survival Probability"
)
```

### Swimmers Plot — Survival by TMB Group

Each bar is a patient. Length represents follow-up time; color indicates vital status. The distribution of follow-up times is similar across TMB groups, reinforcing the non-significant survival finding.

```{r swimmers}
#| fig-height: 10
patient_summary %>%
    filter(!is.na(tmb_group)) %>%
    arrange(tmb_group, overall_survival_days) %>%
    mutate(patient_id = fct_inorder(patient_id)) %>%
    ggplot(aes(x = overall_survival_days, y = patient_id,
               fill = factor(vital_status))) +
    geom_col() +
    facet_wrap(~ tmb_group, scales = "free_y", ncol = 1) +
    scale_fill_manual(
        values = c("0" = "steelblue", "1" = "firebrick"),
        labels = c("0" = "Alive", "1" = "Deceased")
    ) +
    labs(
        x    = "Overall Survival (days)",
        y    = NULL,
        fill = "Status",
        title = "Swimmer Plot by TMB Group"
    ) +
    theme_minimal() +
    theme(axis.text.y = element_blank())
```

---

## Multivariate Cox Regression

### Stage + Age + TMB

Adjusting for stage and age, TMB shows a modest but significant independent association with survival (HR = 1.035, p = 0.013). This effect is not detectable in univariate KM analysis due to small effect size and unbalanced group sizes.

```{r cox-tmb}
cox <- coxph(
    Surv(overall_survival_days, vital_status) ~
        stage_group + age_at_diagnosis_years + tmb,
    data = patient_summary %>% filter(!is.na(stage_group))
)

summary(cox)
```

### Driver Gene Mutations

Binary flags for the six most frequently mutated driver genes were added to the Cox model. After adjusting for stage and age, no individual mutation reaches significance — consistent with these genes being subtype markers rather than independent prognostic factors.

```{r cox-genes}
gene_matrix <- dbGetQuery(con, "
    SELECT
        patient_id,
        MAX(CASE WHEN gene_symbol = 'TP53'   THEN 1 ELSE 0 END) as tp53,
        MAX(CASE WHEN gene_symbol = 'PIK3CA' THEN 1 ELSE 0 END) as pik3ca,
        MAX(CASE WHEN gene_symbol = 'CDH1'   THEN 1 ELSE 0 END) as cdh1,
        MAX(CASE WHEN gene_symbol = 'GATA3'  THEN 1 ELSE 0 END) as gata3,
        MAX(CASE WHEN gene_symbol = 'MAP3K1' THEN 1 ELSE 0 END) as map3k1,
        MAX(CASE WHEN gene_symbol = 'PTEN'   THEN 1 ELSE 0 END) as pten
    FROM staging.mutations
    GROUP BY patient_id
")

survival_genes <- patient_summary %>%
    left_join(gene_matrix, by = "patient_id") %>%
    filter(!is.na(stage_group)) %>%
    mutate(across(c(tp53, pik3ca, cdh1, gata3, map3k1, pten), factor))

cox_genes <- coxph(
    Surv(overall_survival_days, vital_status) ~
        stage_group + age_at_diagnosis_years +
        tp53 + pik3ca + cdh1 + gata3 + map3k1 + pten,
    data = survival_genes
)

tidy(cox_genes, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(!grepl("stage|age", term)) %>%
    mutate(term = toupper(term)) %>%
    ggplot(aes(x = estimate, y = reorder(term, estimate),
               xmin = conf.low, xmax = conf.high)) +
    geom_pointrange() +
    geom_vline(xintercept = 1, linetype = "dashed", color = "firebrick") +
    labs(
        x     = "Hazard Ratio (95% CI)",
        y     = NULL,
        title = "Driver Gene Mutations and Survival",
        subtitle = "Adjusted for stage and age — no gene reaches significance"
    ) +
    theme_minimal()
```

---

```{r save-plots, include=FALSE}
dir.create("../docs_site/docs/assets/img", recursive = TRUE, showWarnings = FALSE)

# KM by stage
p_stage <- ggsurvplot(
    fit_stage,
    data         = patient_summary %>% filter(!is.na(stage_group)),
    pval         = TRUE,
    palette      = c("#2196F3", "#4CAF50", "#FF9800", "#F44336"),
    legend.title = "Stage",
    legend.labs  = c("I", "II", "III", "IV"),
    xlab = "Days", ylab = "Survival Probability"
)
ggsave("../docs_site/docs/assets/img/km_stage.png",
       plot = p_stage$plot, width = 7, height = 5, dpi = 150)

# Forest plot (gene mutations)
p_forest <- tidy(cox_genes, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(!grepl("stage|age", term)) %>%
    mutate(term = toupper(term)) %>%
    ggplot(aes(x = estimate, y = reorder(term, estimate),
               xmin = conf.low, xmax = conf.high)) +
    geom_pointrange() +
    geom_vline(xintercept = 1, linetype = "dashed", color = "firebrick") +
    labs(x = "Hazard Ratio (95% CI)", y = NULL,
         title = "Driver Gene Mutations and Survival",
         subtitle = "Adjusted for stage and age") +
    theme_minimal()
ggsave("../docs_site/docs/assets/img/forest_genes.png",
       plot = p_forest, width = 6, height = 4, dpi = 150)

# Oncoplot — maftools writes directly to device
png("../docs_site/docs/assets/img/oncoplot.png",
    width = 900, height = 600, res = 120)
oncoplot(maf = maf, top = 20)
dev.off()
```

```{r teardown, include=FALSE}
dbDisconnect(con, shutdown = TRUE)
```
